---
title: "Predictive Modeling: BB Fictional Loans Dataset"
subtitle: "Logistic Regression — BadFlagCategory"
author: "Nina Briantseva"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    fig_width: 7
    fig_height: 4.5
    latex_engine: xelatex
geometry: margin=2.5cm
fontsize: 11pt
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{badblue}{HTML}{4b8bbe}
  - \definecolor{badred}{HTML}{e05252}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.pos   = "H"
)
```

```{r packages, include=FALSE}
required <- c("tidyverse", "caret", "pROC", "corrplot", "scales", "knitr",
              "kableExtra", "broom")
for (pkg in required) {
  if (!requireNamespace(pkg, quietly = TRUE))
    install.packages(pkg, repos = "https://cloud.r-project.org")
}
library(tidyverse)
library(caret)
library(pROC)
library(corrplot)
library(scales)
library(knitr)
library(kableExtra)
library(broom)
```

---

\newpage

# Introduction

This report presents a predictive modeling analysis of the **BB Fictional Loans** dataset.
The objective is to build a logistic regression model that predicts whether a loan will
be classified as **bad** (`BadFlagCategory = 1`) or **good** (`BadFlagCategory = 0`).

The workflow covers:

1. Exploratory Data Analysis (EDA)
2. Data cleaning and feature engineering
3. Model training and reduction
4. Evaluation (Accuracy, Precision, Recall, F1, AUC-ROC)
5. Findings and recommendations

---

# Data Overview

```{r load-data}
loans_raw <- read.csv("BB_fictional_loans.csv", stringsAsFactors = FALSE)
```

```{r data-overview}
cat("Rows:", nrow(loans_raw), " | Columns:", ncol(loans_raw))
```

```{r column-types, echo=FALSE}
col_info <- data.frame(
  Column  = names(loans_raw),
  Type    = sapply(loans_raw, class),
  Example = sapply(loans_raw, function(x) as.character(x[1]))
)
kable(col_info, caption = "Column types and example values",
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 9)
```

## Missing Values

```{r missing}
miss <- colSums(is.na(loans_raw))
kable(data.frame(Column = names(miss), Missing = miss),
      caption = "Missing values per column",
      booktabs = TRUE, row.names = FALSE) %>%
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 9)
```

> **Note:** No missing values. However, `Education = "-1"` encodes unknown education level and will be recoded.

---

# Exploratory Data Analysis

## Target Variable Distribution

```{r target-dist, fig.cap="Distribution of BadFlagCategory"}
tbl <- table(loans_raw$BadFlagCategory)
df_tbl <- as.data.frame(tbl) %>%
  mutate(pct = Freq / sum(Freq),
         label = paste0(Freq, "\n(", round(pct * 100, 1), "%)"))

ggplot(df_tbl, aes(x = Var1, y = Freq, fill = Var1)) +
  geom_col(width = 0.45) +
  geom_text(aes(label = label), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("0" = "#4b8bbe", "1" = "#e05252")) +
  scale_y_continuous(labels = comma_format(), expand = expansion(mult = c(0, 0.12))) +
  labs(x = "BadFlagCategory  (0 = Good, 1 = Bad)",
       y = "Count", title = "Loan Quality Distribution") +
  theme_minimal() + theme(legend.position = "none")
```

The dataset is **moderately imbalanced**: 78.3% good loans vs. 21.7% bad loans.

## Numeric Variable Distributions

```{r numeric-boxplots, fig.height=7, fig.cap="Numeric variables by loan quality"}
numeric_vars <- c("Age", "Amount", "ExternalCreditScore",
                  "LoanDuration", "IncomeTotal", "LiabilitiesTotal")

loans_long <- loans_raw %>%
  select(BadFlagCategory, all_of(numeric_vars)) %>%
  pivot_longer(-BadFlagCategory, names_to = "Variable", values_to = "Value")

ggplot(loans_long, aes(x = BadFlagCategory, y = Value, fill = BadFlagCategory)) +
  geom_boxplot(outlier.alpha = 0.15, outlier.size = 0.3) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = c("0" = "#4b8bbe", "1" = "#e05252")) +
  labs(x = "BadFlagCategory", y = NULL) +
  theme_minimal() + theme(legend.position = "none",
                          strip.text = element_text(face = "bold", size = 8))
```

## Bad Loan Rate by Categorical Variables

```{r categorical-rates, fig.height=7, fig.cap="Bad loan rate by categorical predictors"}
cat_vars <- c("NewCreditCustomer", "ExternalPaymentDefaultRemarks",
              "Country", "Gender", "Education")

bad_rates <- map_dfr(cat_vars, function(v) {
  loans_raw %>%
    mutate(is_bad = as.integer(BadFlagCategory == "1")) %>%
    group_by(Category = factor(.data[[v]]), Variable = v) %>%
    summarise(bad_rate = mean(is_bad), n = n(), .groups = "drop")
})

ggplot(bad_rates, aes(x = Category, y = bad_rate, fill = bad_rate)) +
  geom_col() +
  geom_text(aes(label = paste0(round(bad_rate * 100, 0), "%")),
            vjust = -0.4, size = 2.8) +
  facet_wrap(~ Variable, scales = "free_x", ncol = 2) +
  scale_fill_gradient(low = "#aecde8", high = "#e05252") +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Bad Loan Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 7),
        strip.text  = element_text(face = "bold", size = 8),
        legend.position = "none")
```

## Correlation Matrix

```{r corr-matrix, fig.cap="Pearson correlation among numeric variables and target"}
cor_data <- loans_raw %>%
  select(all_of(numeric_vars)) %>%
  mutate(BadFlag = as.integer(loans_raw$BadFlagCategory))

cor_mat <- cor(cor_data, use = "pairwise.complete.obs")
corrplot(cor_mat, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black",
         number.cex = 0.65, mar = c(0, 0, 1, 0))
```

```{r corr-table, echo=FALSE}
corr_df <- as.data.frame(round(cor_mat, 3))
kable(corr_df, caption = "Correlation matrix (numeric variables + target)",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "striped", "scale_down"),
                font_size = 8)
```

> **Key observation:** `IncomeTotal` and `LiabilitiesTotal` have a correlation
of **`r round(cor(loans_raw$IncomeTotal, loans_raw$LiabilitiesTotal), 3)`**.
`ExternalCreditScore` shows the strongest negative correlation with
`BadFlagCategory` (`r round(cor_mat["ExternalCreditScore","BadFlag"], 3)`).

---

# Data Cleaning & Feature Engineering

```{r feature-engineering}
loans <- loans_raw

# Parse dates
loans$LoanDate  <- as.Date(loans$LoanDate)
loans$LoanMonth <- month(loans$LoanDate)

# Recode unknown education
loans$Education[loans$Education == "-1"] <- "Unknown"

# Cap outliers at 99th percentile
cap99 <- function(x) pmin(x, quantile(x, 0.99, na.rm = TRUE))
loans$Amount_c <- cap99(loans$Amount)
loans$Income_c <- cap99(loans$IncomeTotal)
loans$Liab_c   <- cap99(loans$LiabilitiesTotal)
loans$CS_c     <- cap99(loans$ExternalCreditScore)

# Debt-to-income ratio
loans$DebtToIncome <- ifelse(loans$Income_c > 0,
                              loans$Liab_c / loans$Income_c, NA)

# Bin numeric variables
loans$Age_bin <- cut(loans$Age,
  breaks = c(0, 24, 34, 44, 54, 64, Inf),
  labels = c("18-24","25-34","35-44","45-54","55-64","65+"), right = TRUE)

loans$Amount_bin <- cut(loans$Amount_c,
  breaks = quantile(loans$Amount_c, probs = 0:4/4, na.rm = TRUE),
  labels = c("Q1","Q2","Q3","Q4"), include.lowest = TRUE)

loans$CreditScore_bin <- cut(loans$CS_c,
  breaks = c(-Inf, 400, 600, 750, Inf),
  labels = c("Poor","Fair","Good","Excellent"), right = TRUE)

loans$Duration_bin <- cut(loans$LoanDuration,
  breaks = c(0, 12, 24, 36, Inf),
  labels = c("Short","Medium","Long","VeryLong"), right = TRUE)

loans$Income_bin <- cut(loans$Income_c,
  breaks = quantile(loans$Income_c, probs = 0:4/4, na.rm = TRUE),
  labels = c("Low","MidLow","MidHigh","High"), include.lowest = TRUE)

loans$DTI_bin <- cut(loans$DebtToIncome,
  breaks = c(-Inf, 0.2, 0.5, 1.0, Inf),
  labels = c("DTI_Low","DTI_Moderate","DTI_High","DTI_VeryHigh"), right = TRUE)
loans$DTI_bin <- addNA(loans$DTI_bin)
levels(loans$DTI_bin)[is.na(levels(loans$DTI_bin))] <- "DTI_Unknown"

# Factors
loans$BadFlagCategory <- factor(loans$BadFlagCategory, levels = c("0","1"))
loans$NewCreditCustomer <- factor(loans$NewCreditCustomer)
loans$Gender  <- factor(loans$Gender)
loans$Country <- factor(loans$Country)
loans$Education <- factor(loans$Education)
loans$ExternalPaymentDefaultRemarks <- factor(loans$ExternalPaymentDefaultRemarks)
```

## New Features Summary

```{r feat-summary, echo=FALSE}
feat_tbl <- data.frame(
  Feature      = c("LoanMonth","Age_bin","Amount_bin","CreditScore_bin",
                   "Duration_bin","Income_bin","DTI_bin"),
  Description  = c("Month of loan origination (seasonal effect)",
                   "Age in 10-yr bands: 18-24, 25-34, …, 65+",
                   "Loan amount quartiles (outliers capped at p99)",
                   "Credit score bands: Poor / Fair / Good / Excellent",
                   "Duration bands: Short ≤12m, Medium ≤24m, Long ≤36m, VeryLong",
                   "Income quartiles (outliers capped at p99)",
                   "Debt-to-income ratio bands (0.2 / 0.5 / 1.0 thresholds)")
)
kable(feat_tbl, caption = "Engineered features",
      booktabs = TRUE, col.names = c("Feature","Description")) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 9)
```

---

# Model Training

## Train / Test Split

Stratified 70 / 30 split to preserve class balance.

```{r split}
model_vars <- c("BadFlagCategory","NewCreditCustomer","Gender","Country",
                "Education","ExternalPaymentDefaultRemarks",
                "Age_bin","Amount_bin","CreditScore_bin","Duration_bin",
                "Income_bin","DTI_bin","LoanMonth")

model_data <- loans %>%
  select(all_of(model_vars)) %>%
  mutate(LoanMonth = factor(LoanMonth)) %>%
  drop_na()

set.seed(42)
idx   <- createDataPartition(model_data$BadFlagCategory, p = 0.70, list = FALSE)
train <- model_data[idx, ]
test  <- model_data[-idx, ]
```

```{r split-table, echo=FALSE}
split_df <- data.frame(
  Split     = c("Train (70%)", "Test (30%)", "Total"),
  Rows      = c(nrow(train), nrow(test), nrow(model_data)),
  Good_0    = c(sum(train$BadFlagCategory == "0"),
                sum(test$BadFlagCategory  == "0"),
                sum(model_data$BadFlagCategory == "0")),
  Bad_1     = c(sum(train$BadFlagCategory == "1"),
                sum(test$BadFlagCategory  == "1"),
                sum(model_data$BadFlagCategory == "1")),
  Bad_Rate  = round(c(mean(train$BadFlagCategory == "1"),
                      mean(test$BadFlagCategory  == "1"),
                      mean(model_data$BadFlagCategory == "1")), 4)
)
kable(split_df, caption = "Train / test split",
      booktabs = TRUE, col.names = c("Split","Rows","Good (0)","Bad (1)","Bad rate")) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 10)
```

## Full Logistic Regression Model

```{r logit-full}
logit_full <- glm(BadFlagCategory ~ .,
                  data   = train,
                  family = binomial(link = "logit"))
```

```{r full-coef-table, echo=FALSE}
full_coef <- tidy(logit_full) %>%
  mutate(across(c(estimate, std.error, statistic), ~ round(., 4)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001),
         sig = ifelse(as.numeric(gsub("<","",p.value)) < 0.05 |
                        startsWith(p.value, "<"), "*", ""))

kable(full_coef, caption = "Full model coefficients",
      booktabs = TRUE, col.names = c("Term","Estimate","Std.Error","z","p-value","Sig.")) %>%
  kable_styling(latex_options = c("hold_position","striped","scale_down"),
                font_size = 7.5)
```

## Reduced Model (p < 0.05 predictors only)

```{r logit-reduced}
sig_terms <- tidy(logit_full) %>%
  filter(p.value < 0.05, term != "(Intercept)") %>%
  pull(term)

all_preds <- setdiff(names(train), "BadFlagCategory")
sig_vars  <- Filter(function(v) any(startsWith(sig_terms, v)), all_preds)

formula_reduced <- as.formula(
  paste("BadFlagCategory ~", paste(sig_vars, collapse = " + "))
)
logit_reduced <- glm(formula_reduced, data = train,
                     family = binomial(link = "logit"))
```

Significant base variables retained: **`r paste(sig_vars, collapse = ", ")`**

```{r reduced-coef-table, echo=FALSE}
red_coef <- tidy(logit_reduced) %>%
  mutate(across(c(estimate, std.error, statistic), ~ round(., 4)),
         p.value = format.pval(p.value, digits = 3, eps = 0.001))

kable(red_coef, caption = "Reduced model coefficients",
      booktabs = TRUE, col.names = c("Term","Estimate","Std.Error","z","p-value")) %>%
  kable_styling(latex_options = c("hold_position","striped","scale_down"),
                font_size = 7.5)
```

---

# Model Evaluation

```{r eval-func, echo=FALSE}
eval_model <- function(model, test_df, label = "Model", threshold = 0.5) {
  probs  <- predict(model, newdata = test_df, type = "response")
  preds  <- factor(ifelse(probs >= threshold, "1", "0"), levels = c("0","1"))
  actual <- test_df$BadFlagCategory
  cm     <- confusionMatrix(preds, actual, positive = "1")
  roc_obj <- roc(as.numeric(actual) - 1, probs, quiet = TRUE)
  list(cm = cm, roc = roc_obj, auc = auc(roc_obj), probs = probs, label = label)
}

res_full    <- eval_model(logit_full,    test, "Full Model")
res_reduced <- eval_model(logit_reduced, test, "Reduced Model")
```

## Confusion Matrices

```{r cm-full, echo=FALSE}
cm_df_full <- as.data.frame(res_full$cm$table)
kable(res_full$cm$table, caption = "Confusion matrix — Full Model",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 10)
```

```{r cm-reduced, echo=FALSE}
kable(res_reduced$cm$table, caption = "Confusion matrix — Reduced Model",
      booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 10)
```

## Metrics Comparison

```{r metrics-table, echo=FALSE}
get_metrics <- function(res) {
  data.frame(
    Model     = res$label,
    Accuracy  = round(res$cm$overall["Accuracy"],  4),
    Precision = round(res$cm$byClass["Precision"], 4),
    Recall    = round(res$cm$byClass["Recall"],    4),
    F1        = round(res$cm$byClass["F1"],        4),
    AUC_ROC   = round(res$auc,                     4)
  )
}

metrics_tbl <- bind_rows(get_metrics(res_full), get_metrics(res_reduced))
kable(metrics_tbl, caption = "Model performance comparison",
      booktabs = TRUE, row.names = FALSE,
      col.names = c("Model","Accuracy","Precision","Recall","F1","AUC-ROC")) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 10)
```

## ROC Curves

```{r roc-curves, fig.cap="ROC curves — Full vs. Reduced model", fig.height=4.5}
par(mfrow = c(1, 2))

plot(res_full$roc, col = "#4b8bbe", lwd = 2.5, main = "Full Model",
     legacy.axes = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = paste("AUC =", round(res_full$auc, 3)),
       col = "#4b8bbe", lwd = 2)

plot(res_reduced$roc, col = "#e05252", lwd = 2.5, main = "Reduced Model",
     legacy.axes = TRUE)
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = paste("AUC =", round(res_reduced$auc, 3)),
       col = "#e05252", lwd = 2)

par(mfrow = c(1, 1))
```

---

# Key Predictors

```{r coef-plot, fig.cap="Top 20 coefficients — Reduced Logistic Regression", fig.height=6}
coef_df <- tidy(logit_reduced) %>%
  filter(term != "(Intercept)") %>%
  arrange(desc(abs(estimate))) %>%
  head(20)

ggplot(coef_df, aes(x = reorder(term, estimate), y = estimate,
                    fill = estimate > 0)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE"  = "#e05252",
                               "FALSE" = "#4b8bbe"),
                    labels = c("Protective (lowers risk)",
                               "Risk factor (raises risk)")) +
  labs(title  = "Top 20 Coefficients — Reduced Logistic Regression",
       subtitle = "Positive coefficient = higher probability of bad loan",
       x = NULL, y = "Log-odds coefficient", fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.y     = element_text(size = 8))
```

---

# Findings & Recommendations

## Summary Table

```{r findings-table, echo=FALSE}
findings <- data.frame(
  Item  = c("Dataset size","Bad loan rate","Missing values",
            "Full model AUC","Reduced model AUC",
            "Strongest predictor","Class imbalance"),
  Value = c(
    paste0(format(nrow(loans_raw), big.mark = ","), " loans"),
    paste0(round(mean(loans_raw$BadFlagCategory == "1") * 100, 1), "%"),
    "None",
    round(res_full$auc,    4),
    round(res_reduced$auc, 4),
    coef_df$term[1],
    "78.3% Good vs 21.7% Bad"
  )
)
kable(findings, caption = "Summary of findings",
      booktabs = TRUE, col.names = c("Item","Value")) %>%
  kable_styling(latex_options = c("hold_position","striped"), font_size = 10)
```

## Key Conclusions

1. **ExternalCreditScore** and **ExternalPaymentDefaultRemarks** are the strongest
   predictors of loan default — the credit bureau data is highly informative.

2. **NewCreditCustomer** borrowers have a notably different risk profile from returning
   customers, suggesting separate models may improve performance.

3. **Country** and **Income level** contribute meaningfully — the geographic and
   economic context matters.

4. The **reduced model retains** `r round(res_reduced$auc / res_full$auc * 100, 1)`%
   of the full model's discriminatory power with fewer predictors, making it
   preferable for interpretability and production deployment.

5. **Class imbalance** (78/22 split) is moderate. For deployment, consider
   class-weighted training or threshold tuning to optimise Precision/Recall trade-off
   based on the business cost of false negatives vs. false positives.

---

\newpage

# Session Info

```{r session-info, echo=FALSE}
si <- sessionInfo()
cat("R version:", R.version$version.string, "\n")
cat("Platform: ", R.version$platform, "\n\n")
cat("Key packages:\n")
pkgs <- c("tidyverse","caret","pROC","corrplot","knitr","kableExtra")
for (p in pkgs) {
  v <- tryCatch(as.character(packageVersion(p)), error = function(e) "N/A")
  cat(sprintf("  %-15s %s\n", p, v))
}
```
